---
title: "Final Report"
author: "Omima Hassan, Maureen McDonald, Hannah Lo, Yuma Anderson"
date: "12/02/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(tidyverse)
library(mapproj)
library(openintro)
library(stringr)
library(dplyr)
library(ggplot2)
library(plotly)
library(tidyverse)
library(lubridate)
library(stringr)
```

## Introduction {.tabset .tabset-fade}

The average tuition cost for university has risen in the past 20 or so years, and we are interested in comparing the changes in tuition and minimum wage. By analyzing the cost of tuition, the minimum wage, and the cost of living index, we can see how prices have changed since 2000 and where expenses differ in U.S. regions.

### Cost of Living Data 

The first data set we used was from Numbeo. It contained the average cost of living index for each state in the United States in 2018. It has 132 cities with two corresponding values. The first value is the 'cost of living index' which is an indicator of consumer goods like groceries, restaurants, transportation, and utilities. The values are indices relative to New York City. For New York City, all indices are 100. The second value is 'rent index,' and it works the same as the 'cost of living index.' If another city has a rent index of 120, this means on average that cities rent is 20% more expensive than New York City. 

An additional dataset contains the latitude and longitude values for every 28,889 cities. This was only used to enable a spatial visualization of our cost of living dataset.

### Minimum Wage Data

The second dataset we used was from the U.S. Department of Labor. It contained the minimum wage for each state from the years 1969-2018. For minimum wage values lower than the federal,
states are allowed to set lower rates for state specific reasons like number of employees and annual gross sales.

### Tuition Data

The first dataset we used was from the Digest of Education Statistics. It contained the average tuition and fees along with board rates for each type of university in the United States by the year for 1963-2018. 

##

---

```{r, include = FALSE}
cost = read.csv("costliv_cleaned.csv", header = T)
tuition = read.csv("tuitioncost_current.csv", header = T)
```

## Data Cleaning
Initial exploration revealed that some data processing was needed before we could proceed with further visualization. Our data went back to 1969. We only want to look at the last 20 years so we deleted all the data from 1969-1999. 

In the minimum wage date, some states had two minimum wages in one year. This could be because of different parts of the state having a higher minimum then the states, like Minnesota and Minneapolis, or they had a change in the middle of the year. To use this data, we took the average of the two minimum wages if a state had multiple wages for one year. 

The data set for tuition cost was also divided into current and constant costs. Since we wanted to compare the costs throughout the years, we only used current, in order to read the cost in our current value of a dollar.

---

```{r, include=FALSE}
cost = read.csv("costliv_lo.csv", header = T)
convert = read.csv("uscities.csv", header = T)
states = map_data('state')
```

```{r, include=FALSE,warning = FALSE}
cost = cost[-1]
cost = separate(cost, City, into = c("city", "state","country"), sep = ',') 
cost = cost[-3]
cost = separate(cost, state, into = c("delete", "state"), sep = ' ')
cost = cost[-2]
cost$state = abbr2state(cost$state)
convert = convert[,c(1,4,9,10)]
convert$state = convert$state_name
convert = convert[-2]
cost_map = left_join(cost,convert,by = c("city","state"))
```

```{r,include=FALSE}
# template
g = list(
  scope = 'north america',
  showland = TRUE,
  landcolor = toRGB("gray95"),
  subunitcolor = toRGB("gray85"),
  countrycolor = toRGB("gray85"),
  showlakes = TRUE,
  lakecolor = toRGB("white"),
  showsubunits = TRUE,
  showcountries = TRUE,
  resolution = 50,
  projection = list(
    type = 'conic conformal',
    rotation = list(lon = -100)),
  lonaxis = list(
    showgrid = TRUE,
    gridwidth = 0.5,
    range = c(-140, -55),
    dtick = 5),
  lataxis = list(
    showgrid = TRUE,
    gridwidth = 0.5,
    range = c(20, 60),
    dtick = 5))

cost_map %>% group_by(state)%>% summarise("city number"=n())
```

## Cost of Living

To explore areas that have a relatively high cost of living, both the 'Cost of Living Index' and 'Rent Index' were examined. The 'Cost of Living Index' for each major city is mapped below with the size and color of the point an output of the 'Cost of Living Index'. The larger and more yellow the point, the greater the value of 'Cost of Living Index' for that city.

Cost of living cost has two significant areas of spiked prices. Expensive areas can be seen on the east and west coasts.  The states of California and Florida have an especially high concentration of cities with a high 'Cost of Living Index'; however, California and Florida do have the highest proportion of major cities from the dataset with 13 cities being from California and 10 from Florida out of 132 total cities. 


```{r,echo=FALSE,warning=FALSE}
# cost of living index map interactive
p1_int = plot_geo(cost_map, lat = ~lat, lon = ~lng) %>%
  add_markers(text = ~paste(city, state, `Cost.of.Living.Index`), color = ~`Cost.of.Living.Index`, size = ~`Cost.of.Living.Index`, hoverinfo = "text") %>%
  colorbar(title = "Cost of Living Index<br />Relative to<br />New York City<br />as 100") %>%
  layout(title = 'Cost of Living Index for Major US Cities 2018<br />(Hover for city)', geo = g)
p1_int
```
The 'Rent Index' was mapped the same way as the 'Cost of Living Index' and the same geographical pattern was visible; however, the 'Rent Index' contained more extreme values than that of 'Cost of Living.' 

For further investigation, the distribution of both variables were plotted next to each other in a boxplot. Revealing that the rent can be very high in extrema cities in comparison to the median. For example, New York City is 65% more expensive for rent than that of the median. However, the difference of the cost of living in these extreme rent cities compared to the median will be much smaller. For example, New York City is 25% more expensive for cost of living than the median. 

```{r,echo=FALSE}
# boxplot comparing distribution of rent and cost of living index
cost_map_gath = gather(cost_map, key = index, value =value, 3:4)
a = ggplot(cost_map_gath,aes(x=index,y=value))+geom_boxplot(fill='green')+annotate("text", x = 1.5, y = 100, label = " <--- New York City --->") +annotate("text", x = 1.3, y = 90, label = "3 extrema")+annotate("text", x = 2.3, y = 90, label = "8 extrema")+labs(title='Barplot of Cost of Living vs Barplot of Rent') + scale_x_discrete(labels=c("Cost of Living", "Rent"))
ggplotly(a)
```

---

```{r,include=FALSE}
#min formatting
min = read.csv("min_wage_clean.csv", header = T)
min = min[,c(1,20)]
names(min)[1] = 'state'
min$state = str_remove_all(min$state,fixed("\n"))
min$region = tolower(min$state)
names(min)[2] = 'min. wage'
comb_map = min%>%left_join(states,by="region")
```

## Minimum Wage

The minimum wage for each state was mapped. There is a concentration of higher minimum wages on the west and upper east coast. This fits the trend of 'Cost of Living Index' sans the lower east coast. Some of the Southern states (Alabama,Louisiana,etc.) do not have a state minimum wage and adopt the federal. This could cause the expense of the state to not be reflected in their minimum wage. Since minimum wage is integrated with government policy, minimum wage data may be inaccurate to judge overall wage. 

```{r, echo=FALSE}
# min wage map interactive 
min_map = ggplot(comb_map, aes(x=long, y=lat,fill=`min. wage`,text=toupper(region))) + geom_polygon(aes(group = group)) +coord_map() +theme(axis.title.x=element_blank(),axis.title.y=element_blank())+ labs(title='Minimum Wage by State (2018)')
ggplotly(min_map)
```

```{r, include=FALSE, warning = FALSE}
min = read.csv("min_wage_clean.csv", header = T)

# Change empty values to na then combining cleaned results 
X2000= which(min$X2000 == "...")
min$X2000[X2000] = "NA"

X2001= which(min$X2001 == "...")
min$X2001[X2001] = "NA"

X2002= which(min$X2002 == "...")
min$X2002[X2002] = "NA"

X2003= which(min$X2003 == "...")
min$X2003[X2003] = "NA"

X2004= which(min$X2004 == "...")
min$X2004[X2004] = "NA"

X2005= which(min$X2005 == "...")
min$X2005[X2005] = "NA"

X2006= which(min$X2006 == "...")
min$X2006[X2006] = "NA"

X2007= which(min$X2007 == "...")
min$X2007[X2007] = "NA"

X2008= which(min$X2008 == "...")
min$X2008[X2008] = "NA"

X2009= which(min$X2009 == "...")
min$X2009[X2009] = "NA"

X2010= which(min$X2010 == "...")
min$X2010[X2010] = "NA"

X2011= which(min$X2011 == "...")
min$X2011[X2011] = "NA"

X2012= which(min$X2012 == "...")
min$X2010[X2012] = "NA"

X2013= which(min$X2013 == "...")
min$X2013[X2013] = "NA"

X2014= which(min$X2014 == "...")
min$X2014[X2014] = "NA"

X2015= which(min$X2015 == "...")
min$X2015[X2015] = "NA"

X2016= which(min$X2016 == "...")
min$X2016[X2016] = "NA"

X2017= which(min$X2017 == "...")
min$X2017[X2017] = "NA"

X2018= which(min$X2018 == "...")
min$X2018[X2018] = "NA"

X2000= min$X2000=as.numeric(min$X2000)
X2001= min$X2001=as.numeric(min$X2001)
X2002= min$X2002=as.numeric(min$X2002)
X2003= min$X2003=as.numeric(min$X2003)
X2004= min$X2004=as.numeric(min$X2004)
X2005= min$X2005=as.numeric(min$X2005)
X2006= min$X2006=as.numeric(min$X2006)
X2007= min$X2007=as.numeric(min$X2007)
X2008= min$X2008=as.numeric(min$X2008)
X2009= min$X2009=as.numeric(min$X2009)
X2010= min$X2010=as.numeric(min$X2010)
X2011= min$X2011=as.numeric(min$X2011)
X2012= min$X2012=as.numeric(min$X2012)
X2013= min$X2013=as.numeric(min$X2013)
X2014= min$X2014=as.numeric(min$X2014)
X2015= min$X2015=as.numeric(min$X2015)
X2016= min$X2016=as.numeric(min$X2016)
X2017= min$X2017=as.numeric(min$X2017)
X2018= min$X2018=as.numeric(min$X2018)

cleanedMin = na.omit(min)
```

What state had the highest minimum wage?
Create a histogram of state to showing the relationship between state and minimum wage
```{r, echo=FALSE}
min$State.or.other...jurisdiction = str_remove_all(min$State.or.other...jurisdiction, fixed("\n"))
min$State.or.other...jurisdiction = factor(min$State.or.other...jurisdiction, levels = min$State.or.other...jurisdiction[order(min$X2018)])
min_ord = ggplot(min,aes(State.or.other...jurisdiction,X2018))+geom_point()+theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 4))+labs(title = "State Minimum Wage in 2018", x="State", y="Minimum Wage (USD)")
min_ord
```

Then we found the coorelation between minimum wage and tuition. First we had to create a new dataset that contained the mean wages for each year. As seen below, mimimum wage and tuition have a high positive correlation. This means that as tuition increases so does mimimum wage at a similar rate.
```{r, include=FALSE}
tuition$tuition2=vector("numeric", length = 1)
tuition$tuition2=(c('12349','12922','13639','14439','15505','16510','17451','18473','19364','20361','21126','22074','23011','23871','24701','25409','26132','26592','27357'))
tuition$tuition2=as.numeric(tuition$tuition2)
meanwage= c(mean(min$X2000), mean(min$X2001),mean(min$X2002),mean(min$X2003),mean(min$X2004),mean(min$X2005),mean(min$X2006),mean(min$X2007),mean(min$X2008),mean(min$X2009),mean(min$X2010),mean(min$X2011),mean(min$X2012),mean(min$X2013),mean(min$X2014),mean(min$X2015),mean(min$X2016),mean(min$X2017),mean(min$X2018))
```

```{r}
cor(meanwage,tuition$tuition2)
```

---

## Conclusion

Living on the coast is more expensive than inland. There are more cities with extreme rent costs than that of cost of living costs. This could be due to space being a limited commodity while the overall cost of consumer goods are an abundant commodity. Individuals who live in higher cost of living areas can expect to have a higher minimum wage except for the Southern-east coast. 

---

## Individual Contributions {.tabset .tabset-fade}

### Maureen:

I helped construct the background and clean data portions of the final project. I also found the coorelation between minimum wage and tuition. I made graphics to support this in the presentation.

### Hannah:

I scraped all the data for the project. I cleaned the minimum wage data by removing all non numeric values. I explored the cost of living index data for patterns and graphed it spatially, I also explored the rent index data the same way. I found differing distributions between the two and further explore it with histograms of both and one barplot. I created a minimum wage spatial map and minimum wage scatterplot to explore the relationship between minimum wage and cost of living data. I integrated everyone's parts into the final report, debugged, and wrote the conclusion. Finally, I collaborated on the presentation. 

### Omina:

write here

### Yuma:

write here:

##

---

## References

* For the convert data : https://simplemaps.com/data/us-cities
* For the tuition data : https://nces.ed.gov/programs/digest/d18/tables/dt18_330.10.asp
* For the min_ wage data : https://www.dol.gov/whd/state/stateMinWageHis.htm
* For the cost_liv data :
 https://www.numbeo.com/cost-of-living/region_rankings.jsp?title=2018&region=021
* https://plot.ly/r/choropleth-maps/
* https://stackoverflow.com/questions/29037851/how-do-i-plot-us-cities-using-ggplot
* https://garrettgman.github.io/tidying/ 
* https://stackoverflow.com/questions/5411979/state-name-to-abbreviation-in-r
* https://rdrr.io/cran/openintro/man/abbr2state.html
* https://community.rstudio.com/t/how-to-add-my-api-key-into-get-map/15992/5
* https://www.datanovia.com/en/blog/top-r-color-palettes-to-know-for-great-data-visualization/
* https://www.r-graph-gallery.com/13-scatter-plot.html
* https://stackoverflow.com/questions/39123458/how-does-cut-with-breaks-work-in-r
* https://stackoverflow.com/questions/45868625/how-to-reverse-the-default-color-palette-for-ggplot2/45869032#45869032
* https://www.rforexcelusers.com/vlookup-in-r/
* https://ggplot2.tidyverse.org/reference/scale_colour_continuous.html
* https://plot.ly/ggplot2/
* https://plot.ly/ggplot2/interactive-tooltip/
* https://stackoverflow.com/questions/43515112/reversing-default-scale-gradient-ggplot2
* https://plot.ly/r/scatter-plots-on-maps/
* https://stackoverflow.com/questions/18115550/combine-two-or-more-columns-in-a-dataframe-into-a-new-column-with-a-new-name
* https://datascience.stackexchange.com/questions/15589/remove-part-of-string-in-r
* https://sejohnston.com/2012/08/09/a-quick-and-easy-function-to-plot-lm-results-in-r/ 