---
title: "final project"
output: html_document
---
```{r}
library(ggplot2)
library(readr)
library(plyr)
```

```{r}
costliv = read_csv("costliv_cleaned.csv")
minwage = read_csv("min_wage_clean.csv")
tuition = read_csv("tuitioncost_current.csv")
```

# in 2018

```{r}
minwage18 <- minwage[c(1,20)]
tuition18 <- tuition[19, 2:3 ] 
costlivhigh <- subset(costliv, `Cost of Living Index` >= 73,
select=c(X1, city, state, `Cost of Living Index`, `Rent Index`))
costlivlow <- subset(costliv, `Cost of Living Index` < 73,
select=c(X1, city, state, `Cost of Living Index`, `Rent Index`))
```

```{r}
costlivhigh1 <- ggplot(costlivhigh, aes(state, `Cost of Living Index`)) + geom_point(alpha= 0.3) + theme(axis.text.x = element_text(angle = 90))
costlivhigh1

costlivlow1 <- ggplot(costlivlow, aes(state, `Cost of Living Index`)) + geom_point(alpha= 0.3) + theme(axis.text.x = element_text(angle = 90))
costlivlow1
```

```{r}
count(costliv, 'state')
count(costlivhigh, 'state')
count(costlivlow, 'state')
```

## count (descending): California, Florida, Texas, Washington, New York, North Carolina
## highest count states (descending): California, Florida, Washington, New York, Colorado
## lowest count states (descending): Texas, Ohio, Tennessee, Michigan, Missouri, North Carolina, New York, Colorado, 

```{r}
library(tidyverse)
costliv_box = costliv
med = costliv_box %>% group_by(state) %>% summarise(med = median(`Cost of Living Index`))
med1 = med[order(med$med),]
med1$state = factor(med1$state, levels = med1$state)
costlivgen1 <- ggplot(med1, aes(state,med)) + geom_point(color='blue') + theme(axis.text.x = element_text(angle = 90))
costlivgen1
```
High Count states
descending: california, new york, washington, florida, colorado

Low Count States
descending: michigan, ohio, tennessee, missouri, texas

# minimum wage analysis

```{r}
minwagebar <- ggplot(minwage18, aes(`State or other\r\r jurisdiction`, `2018`)) + geom_point(stat = "identity")+ theme(axis.text.x = element_text(angle = 90))
minwagebar
```

```{r}
minwagehigh <- subset(minwage18, `2018` >= 10,
select=c(`State or other\r\r jurisdiction`, `2018`))

minwagemid <- subset(minwage18, `2018` < 10 & `2018` > 8,
select=c(`State or other\r\r jurisdiction`, `2018`))

minwagelow <- subset(minwage18, `2018` < 8,
select=c(`State or other\r\r jurisdiction`, `2018`))
```

```{r}
minwageplothigh <- ggplot(minwagehigh, aes(`State or other\r\r jurisdiction`, `2018`)) + geom_point(stat = "identity")+ theme(axis.text.x = element_text(angle = 90)) + labs(title = "Minimum Wage Over $10/hr by State")

minwageplotmid <- ggplot(minwagemid, aes(`State or other\r\r jurisdiction`, `2018`)) + geom_point(stat = "identity")+ theme(axis.text.x = element_text(angle = 90))+ labs(title = "Minimum Wage Between $10-$8/hr by State")

minwageplotlow <- ggplot(minwagelow, aes(`State or other\r\r jurisdiction`, `2018`)) + geom_point(stat = "identity")+ theme(axis.text.x = element_text(angle = 90)) + labs(title = "Minimum Wage Under $8/hr by State")

minwageplothigh
minwageplotmid
minwageplotlow
```

## highest minimum wages: Washington, Massachusetts, California, Oregon, Arizona & Vermont (tied)

# best state comparison
states of interest: arizona, minnesota, maine, michigan, nebraska, South Dakota, vermont,  ohio, new mexico, new hampshire, nevada
```{r}
costlivminnesota <- subset(costliv, state == 'Minnesota',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivminnesota$`Cost of Living Index`)

costlivvermont <- subset(costliv, state == 'Vermont',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivvermont$`Cost of Living Index`)

costlivnh <- subset(costliv, state == 'New Hampshire',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivnh$`Cost of Living Index`)

costlivmichigan <- subset(costliv, state == 'Michigan',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivmichigan$`Cost of Living Index`)

costlivohio <- subset(costliv, state == 'Ohio',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivohio$`Cost of Living Index`)

costlivnebraska <- subset(costliv, state == 'Nebraska',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivnebraska$`Cost of Living Index`)

costlivnevada <- subset(costliv, state == 'Nevada',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivnevada$`Cost of Living Index`)

costlivsd <- subset(costliv, state == 'South Dakota',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivsd$`Cost of Living Index`)

costlivnm <- subset(costliv, state == 'New Mexico',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivnm$`Cost of Living Index`)

costlivarizona <- subset(costliv, state == 'Arizona',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivarizona$`Cost of Living Index`)
```
descending: minnesota, vermont, new hampshire, michigan, ohio, nebraska, nevada, south dakota, new mexico, arizona

# worst state comparison
states of interest: georgia, indiana, louisiana, penn, south carolina, virginia, cali, hawaii, maryland, new york, washington

```{r}
costlivhawaii <- subset(costliv, state == 'Hawaii',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivhawaii$`Cost of Living Index`)

costlivmaryland <- subset(costliv, state == 'Maryland',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivmaryland$`Cost of Living Index`)

costlivcali <- subset(costliv, state == 'California',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivcali$`Cost of Living Index`)

costlivny <- subset(costliv, state == 'New York',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivny$`Cost of Living Index`)

costlivpenn <- subset(costliv, state == 'Pennsylvania',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivpenn$`Cost of Living Index`)

costlivwashington <- subset(costliv, state == 'Washington',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivwashington$`Cost of Living Index`)

costlivvirginia <- subset(costliv, state == 'Virginia',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivvirginia$`Cost of Living Index`)

costlivindiana <- subset(costliv, state == 'Indiana',
select=c(X1, city, state, `Cost of Living Index`))
median(costlivindiana$`Cost of Living Index`)
```


# WORST STATE: PENNSYLVANIA
```{r}
penncost <- ggplot(costlivpenn, aes(group = state, y = `Cost of Living Index`)) + geom_boxplot()
penncost

median(costlivpenn$`Cost of Living Index`)
```
minimum wage: 7.25


# BEST STATE: ARIZONA
```{r}
arizonacost <- ggplot(costlivarizona, aes(group = state, y = `Cost of Living Index`)) + geom_boxplot()
arizonacost

median(costlivarizona$`Cost of Living Index`)
```
minimum wage of arizona: 10.5

